Re-run, Repeat, Reproduce, Reuse, Replicate:\\Transforming Code into Scientific Contributions

### 2017-07-19 draft.tex
Update Reusable
* renamed all instance of `re-use` into `reuse`

\title{Re-run, Repeat, Reproduce, Re-use, Replicate:\\Transforming Code into Scientific Contributions}
% \title{Run Python, run!}
% \title{The R Quintuplet (R$^5$)}
% \author{Nicolas P. Rougier and Fabien C. Y. Benureau}
\author{
  \textbf{Nicolas P. Rougier}$^{1,2,3,*}$, \textbf{Fabien C. Y. Benureau}$^{1,2,3}$\\
  \begin{minipage}{\textwidth}
    \begin{center}
      \vspace{2mm}
      $^{1}$INRIA Bordeaux Sud-Ouest Talence, France $^{2}$ Institut des
      Maladies Neurodégénératives, Université de Bordeaux, CNRS UMR 5293,
      Bordeaux, France $^{3}$ LaBRI, Université de Bordeaux, Bordeaux INP, CNRS
      UMR 5800, Talence, France\\
      \vspace{2mm}
      $^{*}$Corresponding author:
      \href{mailto:Nicolas.Rougier@inria.fr}{Nicolas.Rougier@inria.fr}         
    \end{center}
  \end{minipage}
}
\date{}

@@

But precisely because it is easy to believe that if a program runs once and gives the expected
results it will do so forever, crucial steps to transform working code into
meaningful scientific contributions are rarely undertaken
\citep{Sandve:2013,Schwab:2000}. Computer science is plagued by replication
problems, in part, precisely because it seems impervious to them. In fact, a
program can fail as a scientific contribution in many different ways for many
different reasons. Borrowing the terms coined by \citeauthor{Goble:2016}
\citep{Goble:2016}, for your program to contribute to Science, you have to make
sure it is re-runnable (R$^1$), repeatable (R$^2$), reproducible (R$^3$),
re-usable (R$^4$) and replicable (R$^5$).\\

@@

Have you ever try to re-run a program you wrote some years ago? Frustratingly, it can often be surprisingly hard. Part of the problem is that technology is evolving at a fast pace and you cannot  know in advance how the system, the software and the libraries your program depends on will evolve. Since you wrote the code, you may have reinstalled or upgraded  your operating system. The compiler, interpreter or set of libraries installed may have been replaced with newer versions. You may find yourself battling with arcane issues of library compatibility---thoroughly orthogonal to your immediate research  goals---to execute again a code \emph{that worked perfectly before}. To be clear, it is impossible to write future-proof code, and the best efforts can be stymied by the smallest change in one of the dependencies. At the same time, modernizing an unmaintained ten-year-old code can reveal itself to be an arduous and expensive undertaking, with each change risking to affect the semantics of the program. Rather trying to predict the future or painstakingly dusting off old code, an often easier and less dangerous solution is to recreate the old execution environment\footnote{To be clear, and although virtual machines are often a great help, this is not always possible. It is, however, \emph{always} more difficult when the original execution environment is not documented.}. For this to happen however, the dependencies in terms of systems, software and libraries must be made clear enough.\\

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 2:}} Random walk (R$^1$)}
# Tested with Python 3
import random

@@

%\clearpage
\section*{Re-usable (R$^{\mathbf 4}$)}

Making your program re-usable means it can be easily used, and modified, by you and other people, inside and outside your lab. Ensuring your program is re-usable is advantageous for a number of reasons.
%
For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference.
%
Furthermore, a source code reflects the results of all the decisions that were made to create it, but not the reasons behind them. In science, where the method and its justification matters as much as the results, those reasons are precious knowledge. For instance, it is not sufficient to observe that a piece of code is doing correct predictions on a learning task, one is interested by the exact algorithm that is employed, and why. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess, …) is a valuable information.
%
Re-usability of course directly benefits others researchers, from your team and outside of it. The easier it is to use your code, the lower is the threshold for other to study, modify and extend it. Scientists constantly face the constraint of time: if a model is available, documented, and can be installed, run and understood all in a few hours, it will be preferred over another that would require weeks to reach the same stage.
%
Moreover, scientists prefer to base their work on results they can verify. A reproducible and re-usable code offers a platform both verified and easy-to-use, fostering the development of derivative works by other researchers. Those derivative works contribute to the impact of your original contribution, and generate citations.
%
Having more people examining and using your code also means that potential errors have a higher chance to be caught. If people start using your program, they will most likely report bugs or malfunctions they encountered while using your program. If you're lucky enough, they might even propose you either bugfixes or improvements, hence improving the overall quality of your software. This process contributes to the long-term reproducibility to the extent people continue to use and maintain the program.\\

However, making your program to be re-usable by others requires a little work. If it is natural, when one writes a program for oneself, to make implicit choices that suit one needs, such implicit choices won't resist the will of a user community. For example, having magic numbers or ill-defined constants in your program might be cumbersome for others. Similarly, it might be acceptable to not have a proper documentation when you're the only user (actually it might not be acceptable because the you now and the you in six months are two different persons), but other users want to know precisely how to use this or that function and what it does exactly.

If we look back at our random walker, we can list all the implicit choices that have been made:
\begin{itemize}
\item Step number has been hard-coded (magic number 10)
\item Step size has been fixed to 1
\item Initial position has been set to 0
\item No possibility of using a user-specified random seed
\item Only one run can be started at once
\item Walker walk along a single dimension
\item ...
\end{itemize}
Here, we won't try to implement all these features at once, but only giving the possibility for a user to implement them if needed.\\

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 5:}} Random walk (R$^4$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C.Y. Benureau
# Release under the BSD 2-clause license
# Tested with Python 3.6 / Numpy 1.12.0 / macOS 10.12.4 / 64 bits architecture
import random
import numpy as np

@@

Having made a software re-usable by others should offer a decent protection
against bugs. Unfortunately, this is not always the case and some recent cases
have dramatically illustrated the tremendous impact a bug can have in Science
\citep{Eklund:2016} or in our every-day life \citep{Durumeric:2014}. This is
why, as explained by Peng et al. \cite{Peng:2006}, {\em the replication of
  important findings by multiple independent investigators is fundamental to
  the accumulation of scientific evidence}. Replicability is the implicit
assumption that any article that does not provide the code source makes: that
the description it provides of the algorithms is sufficiently precise and
complete to re-obtain the results it presents. While every published article
should strive for replicability, even the best efforts may fall short because
an oversight, a typo or a difference between what is evident between the writer
and the reader. In that case, a reproducible code is a crucial backup. Hence,
replicability does not negate the necessity of reproducibility. To illustrate
this, let us consider what could be the textual description of your random
walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model has been implemented using the standard Python random generator initialized with the seed 0. At each iteration, a uniform number between -1 (excluded) and +1 (excluded) was drawn and the sign of the result was used for generating a positive or negative step.}
\end{quotation}
%
If we were to replicate this random walk, for example using Numpy which is a
numerical Python library, from this description, we would be confronted with some
unexpected discrepancies between the original Python version and the replicated
version. The reason is to be found in a subtle difference in the respective
implementations of the RNG between Python and Numpy. As explained in the
\href{https://docs.python.org/3.6/library/random.html}{Python documentation},
{\em Python uses the Mersenne Twister as the core generator. It produces
53-bit precision floats and has a period of 2**19937-1.} Numpy RNG is
also based on the Mersenne Twister generator but there are differences in the
way seed is interpreted when initializing the generator. Fortunately, Both
Numpy and Python offer access to the internal state and we can {\em fix} it to
make the behavior of one RNG to match the other RNG. Here, we want to make the
Numpy RNG to match the Python RNG behavior.But there are also more sublte
problems with this description. If we look more closely at it, we can realize
that nothing is said about the specific case of {\tt 0} when generating a step.
Do whe have to consider {\tt 0} to be a positive or a negative step? Without
further information and without the original code, it is up to the reader to
decide. If this would not make a huge difference for small simulations, it
may nonetheless compromise statistics of the results for long runs.\\

Such perfect quantitative match may seem unnecessary in most cases, yet, in
many replication projects, quantitative comparisons are an objective tests that
the behavior is reproduced. Statistical tests demand more work, and may fall
short of detecting small differences in the algorithms, that may results in
huge difference of behavior in different conditions. In all cases, eyeballing
difference is graphs is always difficult. Therefore, one should always provide
the data behind the graphs to allow such comparisons. Without such explicit
replication effort, it would be hard to discover this faulty behavior and this
is one of the reason why the \href{https://rescience.github.io}{ReScience
  journal} has been created. This journal {\em targets computational research and
encourages the explicit replication of already published research, promoting
new and open-source implementations in order to ensure that the original
research is reproducible}.

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 6:}} Random walk (R$^5$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C.Y. Benureau
# Release under the BSD 2-clause license
# Tested with Python 3.6 / Numpy 1.12.0 / macOS 10.12.4 / 64 bits architecture
import random
import numpy as np

>>>>

\title{Re-run, Repeat, Reproduce, Reuse, Replicate:\\Transforming Code into Scientific Contributions}
% \title{Run Python, run!}
% \title{The R Quintuplet (R$^5$)}
% \author{Nicolas P. Rougier and Fabien C. Y. Benureau}
\author{
  \textbf{Nicolas P. Rougier}$^{1,2,3,*}$, \textbf{Fabien C. Y. Benureau}$^{1,2,3}$\\
  \begin{minipage}{\textwidth}
    \begin{center}
      \vspace{2mm}
      $^{1}$INRIA Bordeaux Sud-Ouest Talence, France $^{2}$ Institut des
      Maladies Neurodégénératives, Université de Bordeaux, CNRS UMR 5293,
      Bordeaux, France $^{3}$ LaBRI, Université de Bordeaux, Bordeaux INP, CNRS
      UMR 5800, Talence, France\\
      \vspace{2mm}
      $^{*}$Corresponding author:
      \href{mailto:Nicolas.Rougier@inria.fr}{Nicolas.Rougier@inria.fr}
    \end{center}
  \end{minipage}
}
\date{}

@@

But precisely because it is easy to believe that if a program runs once and gives the expected
results it will do so forever, crucial steps to transform working code into
meaningful scientific contributions are rarely undertaken
\citep{Sandve:2013,Schwab:2000}. Computer science is plagued by replication
problems, in part, precisely because it seems impervious to them. In fact, a
program can fail as a scientific contribution in many different ways for many
different reasons. Borrowing the terms coined by \citeauthor{Goble:2016}
\citep{Goble:2016}, for your program to contribute to Science, you have to make
sure it is re-runnable (R$^1$), repeatable (R$^2$), reproducible (R$^3$),
reusable (R$^4$) and replicable (R$^5$).\\

@@

Have you ever tried to re-run a program you wrote some years ago? Frustratingly, it can often be surprisingly hard. Part of the problem is that technology is evolving at a fast pace and you cannot  know in advance how the system, the software and the libraries your program depends on will evolve. Since you wrote the code, you may have reinstalled or upgraded  your operating system. The compiler, interpreter or set of libraries installed may have been replaced with newer versions. You may find yourself battling with arcane issues of library compatibility---thoroughly orthogonal to your immediate research  goals---to execute again a code \emph{that worked perfectly before}. To be clear, it is impossible to write future-proof code, and the best efforts can be stymied by the smallest change in one of the dependencies. At the same time, modernizing an unmaintained ten-year-old code can reveal itself to be an arduous and expensive undertaking, with each change risking to affect the semantics of the program. Rather trying to predict the future or painstakingly dusting off old code, an often easier and less dangerous solution is to recreate the old execution environment\footnote{To be clear, and although virtual machines are often a great help, this is not always possible. It is, however, \emph{always} more difficult when the original execution environment is not documented.}. For this to happen however, the dependencies in terms of systems, software and libraries must be made clear enough.\\

@@

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 2:}} Re-runnable random walk (R$^1$)}
# Tested with Python 3
import random

@@

%\clearpage
\section*{Reusable (R$^{\mathbf 4}$)}

Making your program reusable means it can be easily used, and modified, by you and other people, inside and outside your lab. Ensuring your program is reusable is advantageous for a number of reasons.\\

For you, first. Because the you now and the you in two years are two different persons. Details on how to use the code, its limitations, its quirks, may be present to your mind now, but will probably escape you in six months \parencite{Donoho:2009}. Here, comments and documentation can make a significant difference. In fact, a source code reflects the results of all the decisions that were made to create it, but not the reasons behind them. In science, where the method and its justification matters as much as the results, those reasons are precious knowledge. In that context, a comment on how a given parameter was chosen (optimization, experimental data, educated guess), why a library was chosen over another (conceptual or technical?) is valuable information.\\

Reusability of course directly benefits others researchers from your team and outside of it. The easier it is to use your code, the lower is the threshold for other to study, modify and extend it. Scientists constantly face the constraint of time: if a model is available, documented, and can be installed, run and understood all in a few hours, it will be preferred over another that would require weeks to reach the same stage.\\

Moreover, scientists prefer to base their work on results they can verify. A reproducible and reusable code offers a platform both validated and easy-to-use, fostering the development of derivative works by other researchers. Those derivative works contribute to the impact of your original contribution.\\

Having more people examining and using your code also means that potential errors have a higher chance to be caught. If people start using your program, they will most likely report bugs or malfunctions they encountered while using your program. If you're lucky enough, they might even propose you either bugfixes or improvements, hence improving the overall quality of your software. This process contributes to the long-term reproducibility to the extent people continue to use and maintain the program.\\

Despites all this, reusability is often overlooked, and it is not hard to see why. Scientists are rarely trained in software engineering, and reusability can represent an expensive endeavour if undertaken as an afterthought, for little tangible short-term benefits, for a codebase that might, after all, see only a single use. And, in fact, reusability is not as indispensable a requirement as the others one presented here.\\

Yet, some simple measures can tremendously increase reusability, and at the same time strenghten reproducibility or re-runnability.\\

Avoid hardcoded or magic numbers. Magic numbers are numbers present directly in the source code, that do not have a name and therefore can be difficult to interpret semantically. Hardcoded values are variables that cannot be changed through a function argument or a parameter configuration file. To be modified, they involve editing the code, which is cumbersome and error-prone. In the R$^3$ code, the seed and the number of steps are respectively hardcoded and magic.\\

%FIXME add ref
Similarily, code behavior should not have to be changed by commenting/uncommenting code. Modification of the behavior of the code, required when different experiments examine slightly different conditions, should always be explicitely set through parameters accessible to the end-user. This improves reproducibility in two-ways: it allows those conditions to be explicitely tracked in the result files since they contains the parameters used. It also allows to define separate scripts to run or configuration files to load to produce each of the figures of the published paper. With a documention explaining which script or configuration file correspond to which experiment, reproducing the different figures becomes straightforward.\\

% def walk():
%     promenade, total = [], 0
%     # for i in range(20): # Figure 2
%     for i in range(10): # Figure 1
%         step = +1 if random.uniform(-1,+1) > 0 else -1
%         total += step
%         promenade.append(total)
%     return promenade

Documentation, in fact, is one of the most potent tools for reusability. a proper documentation on how to install and run the software often makes the difference for other reasearchers whether they manage to use the software or not.\\

A comment describing what each function does, however evident, can also avoid hours of head-scratching. Great code may need few comments. Scientists, however, rarely qualify as senior developers. Code quality, in science, is often mediocre. Bad, complicated code should be rewritten until is simple enough to explain itself. But realistically, this is hardly ever going to be done: there is simply not enough incentive for it. There, a comment that explains the intention and scientific reasons behind a block of code can tremendously useful.\\

% However, making your program reusable requires a little work. If it is natural, when one writes a program for oneself, to make implicit choices that suit one needs, such implicit choices won't resist the will of a user community. For example, having magic numbers or ill-defined constants in your program might be cumbersome for others. Similarly, it might be acceptable to not have a proper documentation when you're the only user (actually it might not be acceptable because the you now and the you in six months are two different persons), but other users want to know precisely how to use this or that function and what it does exactly. %FIXME: reword
%
% If we look back at our random walker, we can list all the implicit choices that have been made:
% \begin{itemize}
% \item Step number has been hard-coded (magic number 10)
% \item Step size has been fixed to 1
% \item Initial position has been set to 0
% \item No possibility of using a user-specified random seed
% \item Only one run can be started at once
% \item Walker walk along a single dimension
% \item ...
% \end{itemize}
% Here, we won't try to implement all these features at once, but only giving the possibility for a user to implement them if needed.\\

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 5:}} Re-runnable, repeatable, reproducible, reusable random walk (R$^4$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C.Y. Benureau
# Release under the BSD 2-clause license
# Tested with Python 3.6 / Numpy 1.12.0 / macOS 10.12.4 / 64 bits architecture
import random
import numpy as np

@@

Having made a software reusable by others should offer a decent protection
against bugs. Unfortunately, this is not always the case and some recent cases
have dramatically illustrated the tremendous impact a bug can have in Science
\citep{Eklund:2016} or in our every-day life \citep{Durumeric:2014}. This is
why, as explained by Peng et al. \cite{Peng:2006}, {\em the replication of
  important findings by multiple independent investigators is fundamental to
  the accumulation of scientific evidence}. Replicability is the implicit
assumption that any article that does not provide the code source makes: that
the description it provides of the algorithms is sufficiently precise and
complete to re-obtain the results it presents. While every published article
should strive for replicability, even the best efforts may fall short because
an oversight, a typo or a difference between what is evident for the writer
and for the reader. In that case, a reproducible code is a crucial backup. Hence,
replicability does not negate the necessity of reproducibility. To illustrate
this, let us consider what could be the textual description of your random
walker (as it would be written in an article describing it):
%
\begin{quotation}
{\em The model has been implemented using the Mersene Twister generator initialized with the seed 0. At each iteration, a uniform number between -1 (excluded) and +1 (excluded) was drawn and the sign of the result was used for generating a positive or negative step.}
\end{quotation}
%
If we were to replicate this random walk, for example using Numpy which is a
numerical Python library, from this description, we would be confronted with some
unexpected discrepancies between the original Python version and the replicated
version. The reason is to be found in a subtle difference in the respective
implementations of the RNG between Python and Numpy. As explained in the
\href{https://docs.python.org/3.6/library/random.html}{Python documentation},
{\em Python uses the Mersenne Twister as the core generator. It produces
53-bit precision floats and has a period of 2**19937-1.} Numpy RNG is
also based on the Mersenne Twister generator but there are differences in the
way seed is interpreted when initializing the generator. Fortunately, Both
Numpy and Python offer access to the internal state and we can {\em fix} it to
make the behavior of one RNG to match the other RNG. Here, we want to make the
Numpy RNG to match the Python RNG behavior. But there are also more sublte
problems with this description. If we look more closely at it, we can realize
that nothing is said about the specific case of {\tt 0} when generating a step.
Do whe have to consider {\tt 0} to be a positive or a negative step? Without
further information and without the original code, it is up to the reader to
decide. If this would not make a huge difference for small simulations, it
may nonetheless compromise statistics of the results for long runs.\\

Such perfect quantitative match may seem unnecessary in most cases, yet, in
many replication projects, quantitative comparisons are an objective tests that
the behavior is reproduced. Statistical tests demand more work, and may fall
short of detecting small differences in the algorithms, that may results in
huge difference of behavior in different conditions. Without such explicit
replication effort, it would be hard to discover this faulty behavior and this
is one of the reason why the \href{https://rescience.github.io}{ReScience
  journal} has been created. This journal {\em targets computational research and
encourages the explicit replication of already published research, promoting
new and open-source implementations in order to ensure that the original
research is reproducible}.

\noindent \begin{minipage}[c]{\linewidth}
\begin{code}{\textbf{\textsc{Listing 6:}} Replicated random walk (R$^5$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C.Y. Benureau
# Release under the BSD 2-clause license
# Tested with Python 3.6 / Numpy 1.12.0 / macOS 10.12.4 / 64 bits architecture
import random
import numpy as np
